---
title: "Obtaining an Annotated VCF File from Raw Data" 
author: "Alperen Korkmaz" 
format: 
  html:
    theme: cosmo
    toc: true
    toc-title: "Contents"
    code-fold: show
editor: visual 
---

## Introduction to the Bioinformatics Pipeline

We set up our variant calling pipeline in the Linux environment and successfully annotated our VCF file. In this report, we will clean our genomic data, perform exploratory data analysis, and create visualizations to better understand the mutation profiles.

### 1. Loading Required Packages

First, we install and load the necessary libraries for data wrangling and visualization.

```{r}
#| message: false
#| warning: false

library(vcfR)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(knitr)
library(DT) # Added for interactive HTML tables
```

### 2. Importing and Tidying the VCF Data

We import the annotated VCF file generated in our Linux environment. By using the `vcfR2tidy` command, we automatically extract the fixed fields into a structured data frame.

*(Note: Ensure that the VCF file is in the same working directory as this script.)*

```{r}
#| message: false
#| warning: false

vcf <- read.vcfR("F350066296_L01_14_anotation_mutations.vcf")
F350066296 <- vcfR2tidy(vcf)
```

Now, let's take a quick look at our raw data structure using an interactive table.

```{r}
#| warning: false

# Replaced View() with datatable() for HTML rendering
datatable(head(F350066296$fix, 50), 
          options = list(scrollX = TRUE, pageLength = 5),
          caption = "Raw VCF Fixed Fields (First 50 Rows)")
```

### 3. Parsing the Annotation Column (SnpEff)

The genomic data was annotated using SnpEff, but the results are contained within a single `ANN` column separated by pipes (`|`). We need to separate this data to isolate biological consequences. We will also filter for `HIGH` and `MODERATE` impact variants to focus on clinically relevant mutations.

```{r}
#| warning: false

ANNOTATION_MUT <- F350066296$fix |> 
  mutate(ANN_first_result = str_remove(ANN, ",.*")) |> 
  separate(
    col = ANN_first_result,
    into = c("Allele", "Mutation_Type","Degree_of_Impact","Gene","Gene_ID","Feature_Type","Feature_ID","Transcript_Biotype","Rank","HGVS.c","HGVS.p","cDNA_pos","CDS_pos","Protein_pos","Distance", "Warnings"),
    sep = "\\|",
    extra = "drop",
    fill = "right"
  ) |> select(CHROM, POS, REF, ALT ,Gene, Mutation_Type, Degree_of_Impact, HGVS.p, QUAL) |>
  filter(Degree_of_Impact %in% c("HIGH", "MODERATE"))
```

### 4. Preparing the Clinical Database Query Table

To facilitate downstream clinical interpretation, we format the variant locations to be easily searchable in professional databases like **Franklin (Genoox)** and **VarSome**.

```{r}
clinic_search_table <- ANNOTATION_MUT |> 
  mutate(
    genomic_location = paste0(CHROM, ":", POS, " ", REF, ">", ALT),
    HGVS_search_term = paste0(Gene, " ", HGVS.p),
    Varsome_search = paste0(CHROM, ":", POS, " ", REF, ">", ALT),
    Franklin_search = paste0(CHROM, ":", POS, " ", REF, ">", ALT)
  ) |> 
  select(Gene, Degree_of_Impact, Mutation_Type, HGVS.p, Varsome_search, Franklin_search, QUAL) |> 
  filter(Degree_of_Impact %in% c("HIGH", "MODERATE"))

datatable(clinic_search_table, 
          options = list(scrollX = TRUE, pageLength = 5),
          caption = "Actionable Variants for Franklin / VarSome Analysis")
```

### 5. Exploratory Data Visualization

Now we create visualizations to better understand the mutation distribution across our genomic sample.

#### 5.1 Distribution of Mutation Types

This graph primarily shows the distribution of mutation types (e.g., missense, frameshift) among the high and moderate impact variants.

```{r}
clinic_search_table |> 
  filter(!is.na(Mutation_Type)) |> 
  count(Mutation_Type, sort = TRUE) |>
  head(10) |> 
  ggplot(aes(x = reorder(Mutation_Type, n), y = n, fill = Mutation_Type)) +
  geom_col(color = "black") +
  coord_flip() +
  labs(
    title = "Distribution of Mutation Types",
    x = "Mutation Type",
    y = "Frequency"
  ) +
  scale_fill_viridis_d(option = "plasma") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

#### 5.2 The Most Mutated Genes

This graph illustrates the top 10 genes harboring the most significant mutations in the annotated VCF file.

```{r}
clinic_search_table |> 
  filter(!is.na(Gene) & Gene != "") |> 
  count(Gene, sort = TRUE) |> 
  head(10) |> 
  ggplot(aes(x = reorder(Gene, n), y = n, fill = Gene)) +
  geom_col(color = "black", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "The 10 Most Mutated Genes",
    subtitle = "Results from the Annotated VCF File",
    x = "Gene Name",
    y = "Mutation Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", color = "#2c3e50")
  )
```

### 6. Quality Control (QC) Metrics

Finally, this table shows the average and maximum sequencing quality (`QUAL`) of the identified variants, grouped by their degree of biological impact. This step is crucial for verifying the reliability of our variant calls.

```{r}
qual_table <- clinic_search_table %>%
  filter(!is.na(QUAL)) %>%
  mutate(QUAL = as.numeric(QUAL)) %>% # Assuring QUAL is numeric for calculation
  group_by(Degree_of_Impact) %>%
  summarise(
    Variant_Count = n(),
    Average_Quality = round(mean(QUAL, na.rm = TRUE), 2),
    Max_Quality = max(QUAL, na.rm = TRUE)
  ) %>%
  arrange(desc(Average_Quality))

# Printing the summary table
kable(qual_table, caption = "Quality Metrics Based on Impact Degree")
```
